Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-17 bbbq_article.tex
All @bogeert feedback is in the TeX. Thanks Geert! Progress #179

% GvdB: A narrative was missing a bit. Why was this done? 
To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in HNC-I and MHC-II elution studies \richel{refs}, 
for the two topology prediction tools TMHMM and PureseqTM \richel{refs}. 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
% GvdB: Also add a conclusion at the end of the paragraph.
From these findings, we conclude that PureseqTM and TMHMM 
result in prediction of a similar number of TMHs 
and that peptides derived from TMHs have been identified for both MHC-I and -II.

@@

We found a general pattern in the
presentation of epitopes that overlap with a TMH.
For humans, a bacterial and viral pathogen,
as well as for MHC-I and MHC-II,
we found that these are presented more often 
than expected by chance.

@@

It is known that an individual's haplotype
influences the dynamics of a pathogen (see,
for example, \cite{eccleston2017host} for HIV).
The bioinformatics pipeline developed in this
study may be interesting to make in silico
predictions of a SARS-CoV-2 infection.
The most prominent example is the (MHC-I)
haplotype HLA-A*24:02,
which is predicted to present four times as much epitopes derived 
from SARS-CoV-2 TMHs than expected by chance.
This may help the immune system to detect a SARS-CoV-2 infection earlier,
and change the infection's dynamics.

@@

@@

We assume that epitopes 
derived from TMHs are actually presented by MHC-I and MHC-II.
Our analysis (at subsection \ref{subsec:elution_studies}) 
finds that 1.4\% (for MHC-I) and 4.0% (for MHC-II) 
of the epitopes presented 
overlap with a TMH.
It is unsure if this a true finding or a false positive.
One source of noise would stem from imperfect predictions
by the topology prediction software used. 
We have counteracted this by using two different topology prediction
software and confirm these have similar results.
Two biological sources of noise
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion \cite{delong2016pathogenic},
two processes that both result in peptide fragments
that did not originate from genome directly.
Note however, that both elution studies have a
methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy.

@@

% \paragraph{Removal of selenoproteins}

We have removed the 25 selenoproteins found in the human
reference proteome, as both prediction tools struggle with 
selenocysteine, the 21st amino acid incorporated in human proteins.
TMHMM gives a clear error message, forcing us to remove
these selenoproteins from the analysis.
How PureseqTM deals with selenocysteine is unclear,
as it gives no error, nor is it mentioned anywhere,
nor do the authors respond to emails. Due to this, we removed the
selenoproteins from our analysis.

% \paragraph{Topology prediction of TRDD1}

PureseqTM does not predict the topology
of proteins that have less than three amino acids. 
The TRDD1 ('T cell receptor delta diversity 1') protein,
however, is two peptides long. 
We corrected for this by predicting that this protein
is completely cytosolic.

% \paragraph{Bacteria have a different cell membrane}

In this experiment we predicted epitopes that overlap with 
TMHs from a human, bacterial and viral proteome,
would these proteins be expressed in a human host.
Bacteria, however have different cell membranes and cell walls, 
hence different structural requirements for a TMH.
Both topology prediction tools were trained to recognize
human TMHs, thus we cannot be sure that
the transmembrane regions predicted in bacterial proteins
are actually part of an integral membrane protein.
For the purpose of this study, we assume the 
error in topology predictions to be unbiased way towards topology.
In other words: that a bacterial TMH is incorrectly
predicted to be absent just as often as it is incorrectly
predicted to be present elsewhere.

@@

% \paragraph{synonymous mutations}

In our evolutionary experiment, 
we removed variations that were synonymous mutations (i.e.
resulted in the same amino acid, from a different genetic code) 
from our analysis.
There is evidence, however, that these synonymous mutations
do have an effect and may even be evolutionary selected 
for (\cite{hunt2009silent}).
As the possible effect of synonymous mutations is ignored by our
topology prediction software, we do so as well.

% \paragraph{False positives in SNPs}

Regarding the evolutionary conservation of TMHs using SNPs,
again, it is estimated that approximately ten percent
of SNPs is a false positive that result from the methods to determine
a SNP. One example is that sequence variations are incorrectly
detected due to highly similar duplicated sequences \cite{musumeci2010single}.
We assume that these duplications occur as often in TMHs as in
regions around these, hence we expect this not to affect our results.

@@

The NCBI dbSNP database contains millions of SNPs.
As the NCBI API constrains its users to three calls per second,
we had to limited the extent of our analysis
by downloading SNPs to approximately 300 
hours (i.e. the length of a Christmas holiday).

% \paragraph{Selection of SNPs}

We selected to analyze SNPs with the lowest SNP ID first.
This means we have a bias for picking SNPs with
an earlier discovery date.
We expect this bias to be biologically irrelevant
and we find no clear evidence of a bias, as
shown in figure \ref{fig:n_tmhs_per_protein}.
Still, the best way to interpret this figure
is open to discussion.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.

>>>>

% GvdB: A narrative was missing a bit. Why was this done? 
To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution studies \richel{refs}, 
for the two topology prediction tools TMHMM and PureseqTM \richel{refs}. 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
% GvdB: Also add a conclusion at the end of the paragraph.
From these findings, we conclude that PureseqTM and TMHMM 
result in prediction of a similar number of TMHs 
and that peptides derived from TMHs have been identified for both MHC-I and -II.

@@

From this study, two important conclusions can be drawn. 
First, the MHC over-presentation of TMHs is likely a general feature 
and predicted to occur for most haplotypes of both MHC-I and -II 
and for humans as well as bacterial and viral pathogens. 
Second, TMHs are more evolutionary conserved that soluble protein motifs, 
at least in the human proteome.

@@

It is known that an individual's haplotype
influences the dynamics of a pathogen (see,
for example, \cite{eccleston2017host} for HIV) \richel{GvdB: Too vague. How?}.
The bioinformatics pipeline developed in this
study may be interesting to make in silico
predictions of a SARS-CoV-2 infection.
The most prominent example is the (MHC-I)
haplotype HLA-A*24:02,
which is predicted to present four times as much epitopes derived 
from SARS-CoV-2 TMHs than expected by chance.
This may help the immune system to detect a SARS-CoV-2 infection earlier,
and change the infection's dynamics.

@@

\richel{
  GvdB:

  Discussion section needs some work, 
  as here interpretation of the results needs to be given. 
  What is the consequence of the findings? Why is it important.

  Now it is just a list of potential reasons why the results might be flawed. 
  While this is in principle OK, 
  it is also nitpicking as some things affect only a limited number of 
  proteins (like selenomethionines) 
  or you’d expect changes in the other direction (wrong prediction of TMHs for bacteria).

  So offer interpretations. 
  The 2nd part of the introduction would fit here well. I can also help with this.
}

@@

We assume that epitopes 
derived from TMHs are actually presented by MHC-I and MHC-II.
Our analysis (at subsection \ref{subsec:elution_studies}) 
finds that 1.4\% (for MHC-I) and 4.0% (for MHC-II) 
of the epitopes presented 
overlap with a TMH.
It is unsure if this a true finding or a false positive.
One source of noise would stem from imperfect predictions
by the topology prediction software used. 
We have counteracted this by using two different topology prediction
software and confirm these have similar results.
Two biological sources of noise \richel{GvdB: What do you mean by noise?}
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion \cite{delong2016pathogenic},
two processes that both result in peptide fragments
that did not originate from genome directly.
Note however, that both elution studies have a
methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy.

@@

\richel{GvdB: Sounds a but irrelevant to me. Do you really want to mention this in the discussion?}
% \paragraph{Removal of selenoproteins}
%
We have removed the 25 selenoproteins found in the human
reference proteome, as both prediction tools struggle with 
selenocysteine, the 21st amino acid incorporated in human proteins.
TMHMM gives a clear error message, forcing us to remove
these selenoproteins from the analysis.
How PureseqTM deals with selenocysteine is unclear,
as it gives no error, nor is it mentioned anywhere,
nor do the authors respond to emails. Due to this, we removed the
selenoproteins from our analysis.

\richel{GvdB: Also irrelevant. Only 1 protein}
% \paragraph{Topology prediction of TRDD1}
%
PureseqTM does not predict the topology
of proteins that have less than three amino acids. 
The TRDD1 ('T cell receptor delta diversity 1') protein,
however, is two peptides long. 
We corrected for this by predicting that this protein
is completely cytosolic.

\richel{GvdB: Also not really relevant. This is just a minor thing, only relevant if we would see large differences between human and bacteria (but we see not)}
% \paragraph{Bacteria have a different cell membrane}
%
In this experiment we predicted epitopes that overlap with 
TMHs from a human, bacterial and viral proteome,
would these proteins be expressed in a human host.
Bacteria, however have different cell membranes and cell walls, 
hence different structural requirements for a TMH.
Both topology prediction tools were trained to recognize
human TMHs, thus we cannot be sure that
the transmembrane regions predicted in bacterial proteins
are actually part of an integral membrane protein.
For the purpose of this study, we assume the 
error in topology predictions to be unbiased way towards topology.
In other words: that a bacterial TMH is incorrectly
predicted to be absent just as often as it is incorrectly
predicted to be present elsewhere.

@@

\richel{GvdB: This might be relevant, but you need to explain why it is relevant. Thus compare your findings with previous efforts to look at conservation of TMHs}
% \paragraph{synonymous mutations}
%
In our evolutionary experiment, 
we removed variations that were synonymous mutations (i.e.
resulted in the same amino acid, from a different genetic code) 
from our analysis.
There is evidence, however, that these synonymous mutations
do have an effect and may even be evolutionary selected 
for (\cite{hunt2009silent}).
As the possible effect of synonymous mutations is ignored by our
topology prediction software, we do so as well.

\richel{GvdB: Also not really relevant}
% \paragraph{False positives in SNPs}
%
Regarding the evolutionary conservation of TMHs using SNPs,
again, it is estimated that approximately ten percent
of SNPs is a false positive that result from the methods to determine
a SNP. One example is that sequence variations are incorrectly
detected due to highly similar duplicated sequences \cite{musumeci2010single}.
We assume that these duplications occur as often in TMHs as in
regions around these, hence we expect this not to affect our results.

@@

The NCBI dbSNP database contains millions of SNPs.
As the NCBI API constrains its users to three calls per second,
we had to limited the extent of our analysis.

\richel{GvdB: This can go to the methods or results.}
% \paragraph{Selection of SNPs}
%
We selected to analyze SNPs with the lowest SNP ID first.
This means we have a bias for picking SNPs with
an earlier discovery date.
We expect this bias to be biologically irrelevant
and we find no clear evidence of a bias, as
shown in figure \ref{fig:n_tmhs_per_protein}.
Still, the best way to interpret this figure
is open to discussion.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.
\richel{GvdB: Should we include Maxim? He did the alignment of Myc tub. Let’s also ask Frans.}
